### 2010-09-28 Documentation/atropos/atropos.tex
initial updates to atropos.tex
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@649 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate
$n$-class Expectation Maximization (EM) segmentation algorithm which uses 
either parametric or non-parametric finite mixture modelings (FMM) of the class intensities.  
It is capable of incorporating prior labelings, prior probability maps, and/or 
Markov Random Field (MRF) modeling assumptions as spatial priors and geodesically 
propagating segmentation information across neuroanatomy to ensure a 
dense labeling.  Atropos has also been efficiently implemented to handle 
large quantities of possible labelings (in the experimental section, we use
up to 56 classes) with a minimal memory footprint.
In addition to describing the technical and implementation aspects of Atropos, 
this work evaluates its performance on the parcellation problem given two different
training datasets, the LPBA40 dataset from the University of California, Los Angeles
and the NIREP dataset
from University of Iowa.  We evaluate three aspects of the problem: (1) the quality
of automatic cortical extraction from raw data; (2) the quality of
parcellation, given a ground truth cortical extraction; (3) the
quality of automated cortical extraction followed by automated
cortical parcellation.

@@

\section{Introduction}

Establishment of anatomical and functional correspondence, both inter- and intra-subjectively, is
vital in neurodevelopmental and neurological research.  With advancements in image acquisition
 technology, significant investment has been made towards development of
 classification techniques for the cerebrum beginning 
with the seminal work of \cite{Vannier1985} in which NASA satellite image processing software was 
appropriated for statistical classification of the various head tissues in 2-D MR images.  Since
then, a proliferation of techniques has ensued with increasing sophistication.  A chronology of 
this development can be inferred from the relevant reviews of the contemporaneous state-of-the-art
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}.
Concomitant with the distillation of technical matters found in the various papers and reviews are
 important evaluation studies of these segmentation methods \citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

Motivations for such algorithmic development are historically rooted in delineating the white and grey matters and the cerebrospinal fluid.
Following the early statistical approach of \cite{Vannier1985},  other researchers followed 
suit in adopting statistical methods for $n$-tissue anatomical brain segmentation.
Given the ``missing data'' aspect of this problem, a natural optimization strategy is the 
use of the Expectation-Maximization (EM) framework \citep{Dempster1977}.  The work described in \cite{Wells1996} 
was one  of the first to incorporate EM to find a locally optimal solution by iterating between
bias field estimation and tissue segmentation.  A core component of this work was the 
explicit modeling of the tissue intensity values as normal distributions \citep{Cline1990} 
for both 2-D univariate simulated data and T1 coronal images which continues to find utility in contemporary developments.   A secondary component was an 
extended nonparametric approach, also influenced by earlier work \citep{Kikinis1992}, where 
Parzen windowing is used to model the tissue intensity distribution omitting consideration of the underlying bias
field.  Although technically not an EM-based algorithm, the robustness of the latter approach has motivated its continued use even in more recent approaches \citep[e.g.][]{Weisenfeld2009}.

@@

Despite its versatility, we limit our Atropos evaluation concentrating on the latter task of image-based cortical parcellation.  Consistent with our advocacy of open science (not to mention the facilitation of analysis due to accessibility) we perform our evaluation analysis using public data sets.  Organization of this work is as follows:  we first describe the theory behind the various algorithmic components of Atropos while acknowledging that theoretical discussion is available elsewhere, e.g. in previously cited literature.  This is followed by a thorough discussion of implementation which is something not typically discussed but yet is of immense practical utility.  We then showcase Atropos' performance on public data sets in comparison with select algorithms previously mentioned.  Finally, we provide a discussion of our work both in the context of the remainder of this paper and in the larger context of previous and future work.

@@

To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normalized distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent the mean and variance of the $k^{th}$ model.  In the case of multi-modal data, where $y_i$ is a vector quantity, a mulitvariate Gaussian is used defined by the mean vector, $\boldsymbol{\mu}_k$, and covariance matrix, $\Sigma_k$.  An initial estimate for these model parameters are tabulated during the initialization process, e.g. K-means or Otsu thresholding.

A common technique for the non-parametric variant is to define $p_k$ using Parzen windowing of the sample observation histogram of $\mathbf{y}$, i.e.
\begin{align} \label{eq:nonparam}
  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_h^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_h^2} \right)
\end{align}
where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.  $\sigma_h$ is the width of each of the $N_B$ Gaussian kernels.  For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.  The likelihood value is determined by the joint probability given by their product.

In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter.

@@

In order to apply MRF theory,  we assume the random field introduced earlier, $X$, is an MRF and is, therefore, characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., respectively
\begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x}
\end{align}
where $\mathbf{x}$ is any particular labeling configuration on $X$ (in other words, any labeling permutation on $X$ is {\em a priori} possible) and 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right)
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This latter property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Due to these two assumption, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function} and $U(\mathbf{x})$ is the {\em energy function} which can take several forms \citep{Li2001}. In Atropos, as is the case with many other segmentation algorithms of the same family, we choose $U(\mathbf{x})$ such that it is only composed of pairwise interactions between neighboring sites,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta, $\delta_{ij}$, based on the Ising potential or Potts model \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for neighborhood systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

@@

We employ the strategy given in \cite{Ashburner2005} in which the stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$, describing the prior probability that $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the template-based prior information at site $i$.

@@

In order to provide a dense segmentation even in regions where the template-based prior probability is 0, we use the template-based prior label images or probability maps described previously to formulate an optional distance prior probability.
Consider a subset of $\mathcal{I}$, which we denote $R_k$, defined by $\mathrm{Pr}(x_i = l_k) > \mathrm{Pr}(x_i = l_j) \,\,\forall j \in K, j \neq k$ and $\mathrm{Pr}(x_i = l_k) > 0$. 
We also denote the boundary of $R_k$ as $\partial R_k$ and the `peak' as $\rho_{R_k}$ where $i$ and $j$ are sites in $R_k$:
\begin{align}
  \rho_{R_k} = \argmax_{i \in R_k} \left\{ \min_{j \in \partial {R_k}} d\left(i,j\right)\right\}.
\end{align}
In other words, the peak is an interior site (or multiple interior sites) in $R_k$ characterized as having the maximum distance of all sites to any corresponding closest point on the boundary of $R_k$.  We denote this max-min distance as $\Delta_k$.  This allows us to define the distance prior probability for label propagation as follows:
\begin{align}\label{eq:prop}
  \mathrm{Pr}(x_i = l_k) = \left\{
                                               \begin{array}{ll}
                                                \alpha_k \exp \left( -\frac{d(i, \partial R_k)}{\sigma_k} \right) & \text{if } i \text{ is outside the region } R_k \\
                                                \alpha_k + d\left(i, \partial R_k\right)\left(\frac{1.0 - \alpha_k}{\Delta_k}\right) & \text{otherwise}
                                                \end{array}
                                                \right.
\end{align}
The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.  Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.  This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even if the template prior probabilities are zero.

@@

A crucial matter, particularly for propagating labels in the cortical regions, is the choice of distance function.  One of the advantages to surface-based methods is that distances can be calculated along the surface mesh, thus discriminating between proximal cortical surface points and those points which are situated close together due to cortical folding.  In Atropos, we provide two possibilities---(1) a geodesic distance function using a fast marching construction \citep{Osher1988} of the distance function over the image and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.  The differences in label propagation results are illustrated in Fig.~\ref{fig:label_propagation}.

@@

\subsection{Optimization}
Given the likelihood models and prior probabilities whose product is proportional to the posterior probability, the Bayesian segmentation solution is the labeling $\hat{\mathbf{x}}$ which maximizes the posterior probability, i.e.
\begin{align}
  \hat{\mathbf{x}} = \argmax_{\mathbf{x}} \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align}
Similar to its algorithmic predecessors, Atropos employs the EM framework \citep{Dempster1977} which, after initial estimation of the likelihood model parameters, iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf. Eqn. (\ref{eq:likelihood})) 
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
with the maximization of its expectation derived in various publications including \cite{Zhang2001} which yields the optimal mean and variance (or covariance).  The derivation of $\gamma_k$ at each iteration follows that provided in \cite{Ashburner2005}.%
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}

With spatial coherence constraints given by the MRF prior, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with cortical labeling problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the introduction in][for a concise summary of the myriad optimization possibilities]{Marroquin2002} each with their characteristic advantages and disadvantages in terms of computational complexity and accuracy,  Atropos utilizes the well-known greedy strategy known as {\em Iterated Conditional Modes} (ICM)  \citep{Besag1986} which is computationally efficient and provides good performance.

@@

As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
framework to the Insight Toolkit where it can be vetted and further improved by other interested researchers.

@@

The typical processing pipeline begins with an intensity normalization/bias correction
step using the recently developed N4 algorithm \citep{Tustison2010} which is based on the popular nonparametric 
nonuniformity intensity normalization (N3) method introduced in \cite{Sled1998}.    It was previously mentioned that several methods proposed in the literature have taken an integrative view of the segmentation problem by incorporating an intrinsic bias correction step into the actual algorithmic workflow.   The advancements
introduced with the N4 algorithm permit such an adaptive integration with Atropos.  Given recent 
demonstrations suggesting that improved white matter segmentation produces better gain field estimates using N3 
\citep{Boyes2008}, when performing 3-tissue segmentation, we use the posterior probability map of white matter
at the current  iteration as a weighted mask for input to N4.  This is done by setting the `{\ttfamily --weight-image}' option on the N4 command line call (see Listing \ref{listing:n4}) to the posterior probability image corresponding to the white matter produced as output in the Atropos call, i.e. `{\ttfamily Atropos --output}'.

>>>>

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm which uses either
parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  It is capable of incorporating prior labelings,
prior probability maps, and/or Markov Random Field (MRF) modeling
assumptions as spatial priors and geodesically propagating
segmentation information across neuroanatomy to ensure a dense
labeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  In addition
to describing the technical and implementation aspects of Atropos,
this work evaluates its performance on the parcellation problem given
two different training datasets, the BrainWeb dataset from Montreal
Neurological Institute and the Hammers atlas from UCL [FIXME].  We use
the datasets to evaluate three uses of the Atropos segmentation method: (1) K-means
segmentation without use of template data; (2) segmentation with prior
probability maps derived from a group template; (3) N-class EM
segmentation for parcellating the brain into 68 classes based on
training data.  These evaluation studies, combined with illustrative
examples that excercise Atropos options, demonstrate both performance
and wide applicability of this new segmentation tool.

@@

\section{Introduction} Neurological and neurodevelopmental research
rely on anatomical and functional correspondence both within and
between subjects.  As image acquisition technology has advanced,
concomitant investment has been made towards image classification
techniques for the cerebrum beginning with the seminal work of
\cite{Vannier1985} in which NASA satellite image processing software
was appropriated for statistical classification of head tissues in 2-D
MR images.  Since then, a proliferation of techniques has ensued with
increasing sophistication.  The chronology of progress in segmentation
may be tracked through both technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid of the primate brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, a natural optimization strategy is the use of the
Expectation-Maximization (EM) framework \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended nonparametric
approach, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter
approach has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

While Atropos may be applied to multivariate data from arbitrary
modalities, we limit our Atropos evaluation to tissue-classification
in T1 neuroimaging in part due to the abundance of ``gold-standard''
data for this modality.  Consistent with our advocacy of open science
(not to mention the facilitation of analysis due to accessibility) we
also only use publicly available data sets.  Organization of this work
is as follows: we first describe the theory behind the various
algorithmic components of Atropos while acknowledging that theoretical
discussion is available elsewhere, e.g. in previously cited
literature.  This is followed by a thorough discussion of
implementation which is something not typically discussed but yet is
of immense practical utility.  Finally, we provide a discussion of our work both in the
context of the remainder of this paper and in the larger context of
previous and future work.

@@

To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normalized distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  In the case of
multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
covariance matrix, $\Sigma_k$.  An initial estimate for these model
parameters are tabulated during the initialization process,
e.g. K-means or Otsu thresholding.

A common technique for the non-parametric variant is to define $p_k$ using Parzen windowing of the sample observation histogram of $\mathbf{y}$, i.e.
\begin{align} \label{eq:nonparam}
  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_j^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_j^2} \right)
\end{align}
where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.  $\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.  For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.  The likelihood value is determined by the joint probability given by their product.

In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \cite{Ashburner2005}.

@@

In order to apply MRF theory,  we assume the random field introduced earlier, $X$, is an MRF and is, therefore, characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., respectively
\begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x}
\end{align}
where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

@@

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.
{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
  allow a weighted combination defined by the ComputeLocalPosterior
  function in Atropos --- how do we justify / explain this?   clearly,
it's a balance of the global solution (no local prior) and the local
one with spatial prior ... }}

@@

In order to provide a dense segmentation even in regions where the template-based prior probability is 0, we use the template-based prior label images or probability maps described previously to formulate an optional distance prior probability.
Consider a subset of $\mathcal{I}$, which we denote $R_k$, defined by $\mathrm{Pr}(x_i = l_k) > \mathrm{Pr}(x_i = l_j) \,\,\forall j \in K, j \neq k$ and $\mathrm{Pr}(x_i = l_k) > 0$. 
We also denote the boundary of $R_k$ as $\partial R_k$ and the `peak' as $\rho_{R_k}$ where $i$ and $j$ are sites in $R_k$:
\begin{align}
  \rho_{R_k} = \argmax_{i \in R_k} \left\{ \min_{j \in \partial {R_k}} d\left(i,j\right)\right\}.
\end{align}
In other words, the peak is an interior site (or multiple interior sites) in $R_k$ characterized as having the maximum distance of all sites to any corresponding closest point on the boundary of $R_k$.  We denote this max-min distance as $\Delta_k$.  This allows us to define the distance prior probability for label propagation as follows:
\begin{align}\label{eq:prop}
  \mathrm{Pr}(x_i = l_k) = \left\{
                                               \begin{array}{ll}
                                                \alpha_k \exp \left( -\frac{d(i, \partial R_k)}{\sigma_k} \right) & \text{if } i \text{ is outside the region } R_k \\
                                                \alpha_k + d\left(i, \partial R_k\right)\left(\frac{1.0 - \alpha_k}{\Delta_k}\right) & \text{otherwise}
                                                \end{array}
                                                \right.
\end{align}
The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.  Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.  This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even where the template prior probabilities are zero.

@@

A crucial matter, particularly for propagating labels in the cortical
regions, is the choice of distance function.  One of the advantages to
surface-based methods is that distances can be calculated along the
surface mesh, thus discriminating between proximal cortical surface
points and those points which are situated close together due to
cortical folding.  In Atropos, we provide two possibilities---(1) a
geodesic distance function using a fast marching construction
\citep{Osher1988} of the distance function with Atropos's masked
region of interest and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.  The differences in label propagation results are illustrated in Fig.~\ref{fig:label_propagation}.

@@

\subsection{Optimization}
Given the likelihood models and prior probabilities whose product is proportional to the posterior probability, the Bayesian segmentation solution is the labeling $\hat{\mathbf{x}}$ which maximizes the posterior probability, i.e.
\begin{align}
  \hat{\mathbf{x}} = \argmax_{\mathbf{x}} \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align}
Similar to its algorithmic predecessors, Atropos employs the EM
framework \citep{Dempster1977}.  After initial estimation of the
likelihood model parameters, EM iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf. Eqn. (\ref{eq:likelihood})).
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with cortical labeling problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally efficient and provides good performance.

@@

As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
framework to the Insight Toolkit where it can be vetted and further
improved by other interested researchers. 
{\textcolor{red}{FIXME ---- should we save this for a deliverable in
    some grant?}}

@@

The typical processing pipeline begins with an intensity normalization/bias correction
step using the recently developed N4 algorithm \citep{Tustison2010} which is based on the popular nonparametric 
nonuniformity intensity normalization (N3) method introduced in \cite{Sled1998}.    It was previously mentioned that several methods proposed in the literature have taken an integrative view of the segmentation problem by incorporating an intrinsic bias correction step into the actual algorithmic workflow.   The advancements
introduced with the N4 algorithm permit such an adaptive integration with Atropos.  
Recent demonstrations suggest improved white matter segmentation produces better gain field estimates using N3 
\citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we use the posterior probability map of white matter
at the current  iteration as a weighted mask for input to N4.  This is done by setting the `{\ttfamily --weight-image}' option on the N4 command line call (see Listing \ref{listing:n4}) to the posterior probability image corresponding to the white matter produced as output in the Atropos call, i.e. `{\ttfamily Atropos --output}'.
